Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2022-02-18 paper/paper_TE_bilingual_vocabulary_model.tex
updated paper

Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English--French bilingual child must use the word ``apple'' when speaking English, and the word \emph{``pomme''} when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children's first words (e.g., David \& Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual's developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak \& Byers-Heinlein, 2019). That is, upon hearing the English word ``apple,'' the corresponding French word \emph{``pomme''} is more easily activated in bilinguals' minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a ``singlet''; Umbel et al., 1992) before they can learn its translation equivalent (i.e., a second label for a referent, in a different language, learned following a singlet)\footnote{We note that in some cases a singlet may not have a close translation equivalent, for example the recently popular Danish word ``hygge,'' which refers to a type of cozy contentment and enjoyment of simplicity. Moreover, even when a word that could be considered a translation equivalent exists, adults' concepts may not be congruent across languages (e.g., Malt \& Majid, 2013). However, in early development, vocabulary words tend to be concrete, and evidence suggests that concepts are largely shared across languages (Storms et al., 2015; White et al., 2020). Thus, we make the assumption that most early words have a translation equivalent and that these are fairly congruent for the age range of focus in the current paper.}. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents and singlets are learned similarly (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals' vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

Taking into account the contributions of language exposure and developmental level to bilingual children's vocabulary acquisition, we put forward the Bilingual Vocabulary Model. This model proposes that the number of translation equivalents that bilingual children produce is a function of vocabulary learning in each language, in the context of the number of potentially learnable words given the children's developmental level. We formalize learning a translation equivalent pair as the joint probability of learning each of the words in the pair. This provides a straightforward empirical test of different theoretical accounts of translation equivalent learning, by asking whether or not the probability of knowing a word is independent of knowing its translation equivalent. The logic is similar to that of the familiar chi-squared test for independence, where the independence of two events from the same population is tested as the probability of their intersection computed by multiplying the probability of each individual event: P(A and B) = P(A) × P(B) if A and B are independent\footnote{To test the independence of two events from the same population, as an example, we might ask whether Psychology majors are more likely to be left-handed. Imagine a college of 1000 students. If 100/1000 (or 1/10) students are left-handed, and 200/1000 (or 1/5) students are psychology majors, then if these variables are independent we expect a proportion of 1/10 × 1/5 = 1/50 students to be left-handed psychology majors. To determine the number we expect to observe in the college, we multiply 1/50 × 1000 = 20 students. When we compare this expected number to the actual number of left-handed students, there are three possible outcomes. First, we may observe many more than 20 left-handed psychology students at the college (say 100 students), which suggests that being left-handed increases the probability of majoring in psychology. Or, second, we may observe many fewer than 20 left-handed psychology students at the college (say 5 students), which suggests that being left-handed decreases the probability of majoring in psychology (in this example by a factor of 1/4). Finally, if left-handedness and majoring in psychology are independent, we can predict the number of left-handed psychology students by multiplying the observed number of left-handed students (100) by the observed number of psychology students (200), and dividing by the total population of the college (1000), so for example 100×200/1000 = 20. Thus, comparing expected and observed numbers can inform us about the independence of the underlying phenomena.}. In the next paragraphs, we define each of the model parameters in detail, and these are also summarized in Table 1.

The model takes four main parameters: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS \textgreater{} 1) or against (BIAS \textless{} 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI. For convenience, consider the effect of setting LEARNABLE to 600, as a round number (the actual number of CDI items is usually slightly higher than 600, depending on the language of the adaptation). However, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019). That is, very young children would not be expected to produce many of the advanced words on the CDI, such as ``lawn mower,'' ``sidewalk,'' or ``vitamins.'' Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child's developmental level, which could be approximated by their age. Since there are no CDI norms for French---English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable --- if imperfect --- estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie's age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age --- a simplifying choice that we will return to in the discussion section.

Using the mathematical concept of independence, we can then quantify the number of translation equivalents (TE) expected given children's vocabulary sizes in the dominant (DOM) and non-dominant (NONDOM) languages, as well as the number of potentially learnable words (LEARNABLE). If dominant-language and non-dominant-language words are learned independently from each other, we multiply DOM × NONDOM (the number of words known in the dominant and non-dominant language respectively), and divide by the total population of learnable words in one language (LEARNABLE) --- which is the possible number of words that could overlap across both languages --- to predict the number of translation equivalents. We further introduce the bias parameter (BIAS), which allows us to examine whether translation equivalent learning is best described by the Avoidance, Preference, or Neutral Account. Adding this parameter, translation equivalents can be derived from TE = BIAS × (DOM × NONDOM)/LEARNABLE. For the Avoidance Account, BIAS will be less than 1, meaning that translation equivalents are less easily learned than singlets; for the Preference Account, BIAS will be greater than 1, meaning that translation equivalents are more easily learned than singlets; for the Neutral Account, BIAS is exactly 1 (i.e., the model is unbiased with respect to whether translation equivalents are more difficult or easier to acquire than singlets). Going back to the example of 18-month-old Jamie, we would set the denominator at 240 which is the number of potentially learnable words at 18 months. If translation equivalents are half as easy to learn as singlets (following the Avoidance Account), we would expect Jamie to produce .5×(50×20/240) = 2.1 translation equivalents. Conversely, if translation equivalents are twice as easy to learn as singlets (following the Preference Account), we would expect Jamie to produce 2×(50×20/240) = 8.3 translation equivalents. Under the Neutral Account, we would expect Jamie to produce 1×(50×20/240) = 4.2 translation equivalents.

@@

Word vocabulary (WORD; sometimes called total productive vocabulary) is the total number of words that a child produces across the two languages, calculated as the sum of the dominant vocabulary (DOM) and non-dominant vocabulary (NONDOM). Concept vocabulary (CONCEPT; sometimes called total conceptual vocabulary) is the number of concepts that are lexicalized by the child --- that is, the total number of concepts that are lexicalized in either language. This can be calculated by subtracting the number of translation equivalents (TE) from the word vocabulary (WORD). Finally, we can also calculate singlets that are produced in each language, that is words for which the child does not yet produce a translation equivalent. Singlets in the dominant language (DOM-SINGLET) can be calculated by subtracting translation equivalents (TE) from dominant-language vocabulary (DOM); singlets in the non-dominant language (NONDOM-SINGLET) can be calculated by subtracting translation equivalents (TE) from non-dominant language vocabulary (NONDOM). It is also possible to decompose children's word vocabulary (WORD) into the sum of TE, DOM-SINGLET, and NONDOM-SINGLET.

@@

Study 1 provides a computational implementation of the Bilingual Vocabulary Model outlined in the introduction, which we use to simulate different scenarios to examine the effect of vocabulary sizes and developmental variables on translation equivalent learning. Note that usually only three values are necessary to calculate all the other variables (see Table 1). Most commonly, we can calculate other variables based on the total number of learnable words (LEARNABLE) together with either the words known in each language (DOM and NONDOM) or word vocabulary plus balance (WORD and BALANCE) which allow us to compute DOM and NONDOM. It is also possible to calculate other variables based on the total number of learnable words (LEARNABLE) with balance and words known in either language (BALANCE and DOM or BALANCE and NONDOM).

We present three simulations generated by creating data points for hypothetical children by varying different parameters of the Bilingual Vocabulary Model, in order to explore expected patterns of translation equivalent learning. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except those that violated constraints in the model (e.g., NONDOM could not be larger than DOM). In the first simulation, we examined how translation equivalent learning relates to vocabulary balance (BALANCE), as well as different metrics of vocabulary size, including dominant-language vocabulary (DOM), non-dominant language vocabulary (NONDOM), and word vocabulary (WORD). In the second simulation, we explored relations between translation equivalents (TE), balance (BALANCE), and learnable words (LEARNABLE). In the first two simulations, the BIAS parameter was held constant at 1 (Neutral Account); in the third simulation, we varied the BIAS parameter to compare translation equivalent learning under the Avoidance, Preference, and Neutral Accounts. A summary of the parameter values used in each simulation is provided in Table 2.

@@

In Simulation 1, we first illustrate the relationships between different variables in the model by simulating three hypothetical children who are at the same developmental level and thus have the same number of potentially learnable words (LEARNABLE), but with different word vocabularies (WORD) and BALANCE. For convenience, we set LEARNABLE = 600 in this example, which roughly corresponds to what is expected for an English--French 27-month-old bilingual child (i.e., the most verbal 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and the most verbal 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages; Frank et al., 2016). We set BIAS to 1, meaning that in these examples translation equivalents and singlets are equally easy to learn.

Our first simulated infant, Annie (small vocabulary, unbalanced exposure), produces 270 words in the dominant language and 30 words in her non-dominant language. She has a word vocabulary of 300, and a balance score of .10 (10\% of her words are in the non-dominant language). Based on the formula TE = DOM×NONDOM/LEARNABLE (we drop BIAS from the formula since it is 1 here) and as seen in Table 3, Annie is expected to produce 13.5 translation equivalents. Our second infant, Bernie (small vocabulary, balanced exposure), produces 180 dominant-language words and 120 non-dominant language words. Like Annie, he has a word vocabulary of 300, but he has a higher balance score of .40 (40\% of his words are in the non-dominant language). Based on our formula, we expect Bernie to produce 36 translation equivalents. Comparing Annie and Bernie, two children who produce the same total number of word vocabulary (i.e., WORD is held constant), the child with the more balanced bilingual vocabulary (Bernie) is expected to produce more translation equivalents. Like Bernie, our third infant Charlie also has a balanced vocabulary, but has a larger word vocabulary (WORD), producing 540 words in the dominant language (DOM) and 360 in the non-dominant language (NONDOM), for a total of 900 words (WORD), and thus BALANCE = .40. Based on our formula for Simulation 1, we expect Charlie to produce 324 translation equivalents (TE). Infants Bernie and Charlie illustrate that for two children equal in BALANCE, the child with larger word vocabulary (WORD) is expected to produce more translation equivalents (TE). Other vocabulary metrics are calculated for each hypothetical child as described in Table 3.

We then broadened this simulation to the more general case and examined patterns of translation equivalent learning, where simulated children had the capacity to learn 600 words (LEARNABLE held constant at 600), and their vocabulary size in each language (DOM and NONDOM) varied. BIAS remained constant at 1. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children's word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)). In Figure 1, we plotted TE knowledge as a function of DOM, NONDOM, and WORD at different levels of BALANCE. Across all three Panels (1A, 1B, and 1C), simulated children with the most balanced vocabulary consistently produced more translation equivalents than other children. Moreover, Panels 1A and 1C show that, as DOM (dominant language words) and WORD (word vocabulary) increased, TE (translation equivalents) also increased regardless of BALANCE. Interestingly, Panel 1B shows that NONDOM and TE were extremely tightly coupled. In sum, we observed three important patterns, which served as Prediction Set 1 from the Bilingual Vocabulary Model for Study 2:

\caption{\label{tab:table3}Table 3. Examples for Simulation 1 of three simulated children with different hypothetical word vocabularies (WORD) and vocabulary balances (BALANCE), where the number of learnable words (LEARNABLE) = 600 and BIAS = 1. Vocabulary values are rounded to the nearest integer.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{100px}>{\raggedright\arraybackslash}p{140px}>{\raggedright\arraybackslash}p{140px}>{\raggedright\arraybackslash}p{60px}>{\raggedright\arraybackslash}p{60px}>{\raggedright\arraybackslash}p{60px}}
\toprule
Variable & Definition & Calculation & Infant Anne (small vocabulary, unbalanced) & Infant Bernie (small vocabulary, balanced) & Infant Charlie (large vocabulary, balanced)\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Main Parameters}}\\
\hspace{1em}BIAS & Bias parameter &  & 1 & 1 & 1\\
\hspace{1em}LEARNABLE & Learnable words in each language &  & 600 & 600 & 600\\
\hspace{1em}DOM & Words produced in the dominant language &  & 270 & 180 & 540\\
\hspace{1em}NONDOM & Words produced in the non-dominant language &  & 30 & 120 & 360\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Derived Parameters}}\\
\hspace{1em}WORD & Word vocabulary (or total vocabulary size) & DOM + NONDOM & 300 & 300 & 900\\
\hspace{1em}BALANCE & Vocabulary balance & NONDOM / (DOM + NONDOM) & 0 & 0 & 0\\
\hspace{1em}TE & Translation equivalents produced & DOM × NONDOM / LEARNABLE & 14 & 36 & 324\\
\hspace{1em}CONCEPT & Concept vocabulary (or total conceptual vocabulary size) & WORD - TE & 286 & 264 & 576\\
\hspace{1em}DOM-SINGLET & Singlets in dominant language & DOM - TE & 256 & 144 & 216\\
\hspace{1em}NONDOM-SINGLET & Singlets in non-dominant language & NONDOM - TE & 16 & 84 & 36\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\begin{itemize}
\tightlist
\item
  Prediction 1a: Children with more balanced vocabularies (BALANCE) will produce more translation equivalents (TE) (Figure 1, Panel 1A).
\item
  Prediction 1b: Children who produce more total words (WORD) or more dominant-language words (DOM) will produce more translation equivalents (TE) (Figure 1, Panels 1A and 1B).
\item
  Prediction 1c: Children who produce more non-dominant language words (NONDOM) will produce more translation equivalents (TE). Unlike for WORD and DOM this will not interact with BALANCE; instead, non-dominant vocabulary size will be an almost perfect predictor of translation equivalent knowledge (Figure 1, Panel 1C).
\end{itemize}

@@

In our previous simulation, we assumed that each simulated child was at the same developmental level and had the capacity to learn up to 600 words in each language (i.e., LEARNABLE held constant at 600). As laid out in the introduction, under the Bilingual Vocabulary Model, the learnability of different words changes with a child's developmental level, where LEARNABLE increases as a child grows older. Therefore, in Simulation 2, we looked at the expected patterns of translation equivalent learning across varying levels of LEARNABLE (i.e., the number of learnable words in each language as developmental level changes). Additionally, we further examined vocabulary composition by computing the number of singlets in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) language. BIAS was once again kept constant at 1.

Translation equivalent knowledge was simulated across children at six developmental levels (number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of CONCEPT, TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. As shown in Panel 1A, simulated children at a later developmental level had larger concept vocabularies (CONCEPT). In Panel 1B, we continued to observe a pattern reported in prediction 1a, whereby simulated children with more balanced vocabularies produced more translation equivalents (TE). Moreover, regardless of balance, simulated children at later developmental levels (i.e., older children with more potentially LEARNABLE words) acquired more translation equivalents (TE). Panels 1C and 1D show that simulated children at later developmental levels also produced more singlets in the dominant and non-dominant languages, and that more balanced children produced more singlets in their non-dominant language, but fewer singlets in their dominant language than less balanced children. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will

@@

\hypertarget{simulation-3-bias-towards-or-against-translation-equivalent-learning-compared-to-singlets}{%
\subsection{1.3 Simulation 3: Bias towards or against translation equivalent learning compared to singlets}\label{simulation-3-bias-towards-or-against-translation-equivalent-learning-compared-to-singlets}}

>>>>

Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English--French bilingual child must use the word ``apple'' when speaking English, and the word \emph{``pomme''} when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children's first words (e.g., David \& Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual's developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak \& Byers-Heinlein, 2019). That is, upon hearing the English word ``apple,'' the corresponding French word \emph{``pomme''} is more easily activated in bilinguals' minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a ``singlet''; Umbel et al., 1992) before they can learn its translation equivalent (i.e., a second label for a referent, in a different language, learned following a singlet)\footnote{We note that in some cases a singlet may not have a close translation equivalent, for example the recently popular Danish word \emph{``hygge''}, which refers to a type of cozy contentment and enjoyment of simplicity. Moreover, even when a word that could be considered a translation equivalent exists, adults' concepts may not be congruent across languages (e.g., Malt \& Majid, 2013). However, in early development, vocabulary words tend to be concrete, and evidence suggests that concepts are largely shared across languages (Storms et al., 2015; White et al., 2020). Thus, we make the assumption that most early words have a translation equivalent and that these are fairly congruent for the age range of focus in the current paper.}. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents and singlets are learned similarly (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals' vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

Taking into account the contributions of language exposure and developmental level to bilingual children's vocabulary acquisition, we put forward the Bilingual Vocabulary Model. This model proposes that the number of translation equivalents that bilingual children produce is a function of vocabulary learning in each language, in the context of the number of potentially learnable words given the children's developmental level. We formalize learning a translation equivalent pair as the joint probability of learning each of the words in the pair. This provides a straightforward empirical test of different theoretical accounts of translation equivalent learning, by asking whether or not the probability of knowing a word is independent of knowing its translation equivalent. The logic is similar to that of the familiar chi-squared test for independence, where the independence of two events from the same population is tested as the probability of their intersection computed by multiplying the probability of each individual event: P(A and B) = P(A) × P(B) if A and B are independent\footnote{To test the independence of two events from the same population, as an example, we might ask whether psychology majors are more likely to be left-handed. Imagine a college of 1000 students. If 100/1000 (or 1/10) students are left-handed, and 200/1000 (or 1/5) students are psychology majors, then if these variables are independent we expect a proportion of 1/10 × 1/5 = 1/50 students to be left-handed psychology majors. To determine the number we expect to observe in the college, we multiply 1/50 × 1000 = 20 students. When we compare this expected number to the actual number of left-handed students, there are three possible outcomes. First, we may observe many more than 20 left-handed psychology students at the college (say 100 students), which suggests that being left-handed increases the probability of majoring in psychology. Or, second, we may observe many fewer than 20 left-handed psychology students at the college (say 5 students), which suggests that being left-handed decreases the probability of majoring in psychology (in this example by a factor of 1/4). Finally, if left-handedness and majoring in psychology are independent, we can predict the number of left-handed psychology students by multiplying the observed number of left-handed students (100) by the observed number of psychology students (200), and dividing by the total population of the college (1000), so for example 100×200/1000 = 20. Thus, comparing expected and observed numbers can inform us about the independence of the underlying phenomena.}. In the next paragraphs, we define each of the model parameters in detail, and these are also summarized in Table 1.

The model takes four main parameters as predictors: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS \textgreater{} 1) or against (BIAS \textless{} 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI, which is typically 600--700 depending on the language of the adaptation. Yet, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019). That is, very young children would not be expected to produce many of the advanced words on the CDI, such as ``lawn mower,'' ``sidewalk,'' or ``vitamins.'' Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child's developmental level, which could be approximated by their age. Since there are no CDI norms for French--English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable --- if imperfect --- estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie's age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age --- a simplifying choice that we will return to in the discussion section.

Using the mathematical concept of independence, we can then quantify the number of translation equivalents (TE) expected given children's vocabulary sizes in the dominant (DOM) and non-dominant (NONDOM) languages, as well as the number of potentially learnable words (LEARNABLE). If dominant-language and non-dominant-language words are learned independently from each other, we multiply DOM × NONDOM (the number of words known in the dominant and non-dominant language respectively), and divide by the total number of learnable words in one language (LEARNABLE) --- which is the possible number of words that could overlap across both languages --- to predict the number of translation equivalents. We further introduce the bias parameter (BIAS), which allows us to examine whether translation equivalent learning is best described by the Avoidance, Preference, or Neutral Account. Adding this parameter, translation equivalents can be derived from TE = BIAS × (DOM × NONDOM)/LEARNABLE. For the Avoidance Account, BIAS will be less than 1, meaning that translation equivalents are less easily learned than singlets; for the Preference Account, BIAS will be greater than 1, meaning that translation equivalents are more easily learned than singlets; for the Neutral Account, BIAS is exactly 1 (i.e., the model is unbiased with respect to whether translation equivalents are more difficult or easier to acquire than singlets). Going back to the example of 18-month-old Jamie, we would set the denominator at 240 which is the number of potentially learnable words at 18 months. If translation equivalents are half as easy to learn as singlets (following the Avoidance Account), we would expect Jamie to produce .5×(50×20/240) = 2.1 translation equivalents. Conversely, if translation equivalents are twice as easy to learn as singlets (following the Preference Account), we would expect Jamie to produce 2×(50×20/240) = 8.3 translation equivalents. Under the Neutral Account, we would expect Jamie to produce 1×(50×20/240) = 4.2 translation equivalents.

@@

Word vocabulary (WORD; sometimes called total vocabulary) is the total number of words that a child produces across the two languages, calculated as the sum of the dominant vocabulary (DOM) and non-dominant vocabulary (NONDOM). Concept vocabulary (CONCEPT; sometimes called total conceptual vocabulary) is the number of concepts that are lexicalized by the child --- that is, the total number of concepts that are lexicalized in either language. This can be calculated by subtracting the number of translation equivalents (TE) from the word vocabulary (WORD). Finally, we can also calculate singlets that are produced in each language, that is words for which the child does not yet produce a translation equivalent. Singlets in the dominant language (DOM-SINGLET) can be calculated by subtracting translation equivalents (TE) from dominant-language vocabulary (DOM); singlets in the non-dominant language (NONDOM-SINGLET) can be calculated by subtracting translation equivalents (TE) from non-dominant language vocabulary (NONDOM). It is also possible to decompose children's word vocabulary (WORD) into the sum of TE, DOM-SINGLET, and NONDOM-SINGLET.

@@

Study 1 provides a computational implementation of the Bilingual Vocabulary Model outlined in the introduction, which we use to simulate different scenarios to examine the effect of vocabulary sizes and developmental variables on translation equivalent learning. Simulations were conducted by holding constant or varying the main parameters across a set of possible values, and computing expected translation equivalents and derived parameters from these values. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except those that violated constraints in the model (e.g., NONDOM could not be larger than DOM).

For each simulation, we generated predictions for expected patterns in bilingual children's vocabulaires. Several previous studies in the literature have examined relations between translation equivalent knowledge and our main parameters including age or developmental level (which determines LEARNABLE; David \& Wei, 2008; Legacy et al., 2016), DOM (Legacy et al., 2016; Poulin-Dubois et al., 2013), and NON-DOM (Legacy et al., 2016). However, others have examined derived parameters such as BALANCE (David \& Wei, 2008; Legacy et al., 2016; Montanari, 2010; Pearson et al., 1995; 1997), WORD (Legacy et al., 2016; Montanari, 2010), CONCEPT (Pearson et al., 1993), DOM-SINGLET and NONDOM-SINGLET (De Houwer et al., 2006; Pearson et al., 1993; 1995). To facilitate comparison with the extant literature where main parameters are not always reported, we generated predictions expressed both in terms of main parameters and derived parameters. This provides alternate ways of testing the model, depending on which parameters are available in the dataset. Note that predictions from the main parameters and those from derived parameters do not provide independent tests of our model for the same dataset, as they can be algebraically calculated from one another. As an analogy, a child's age is not independent from their birth date and today's date, yet we might make reference to their age in some situations and their birth date in others. Similarly, elucidating relations among both main and derived parameters provides a more complete description of expected patterns in bilingual vocabulary development.

In the first simulation, we examined relations between translation equivalents (TE), dominant-language vocabulary (DOM), non-dominant language vocabulary (NONDOM), word vocabulary (WORD), and vocabulary balance (BALANCE), while keeping the number of learnable words (LEARNABLE) constant. In the second simulation, we varied the number of learnable words (LEARNABLE), and examined expected number of translation equivalents (TE), singlets in each language (DOM-SINGLET and NONDOM-SINGLET), and concept vocabulary (CONCEPT). In both of the first two simulations, the BIAS parameter was held constant at 1 (Neutral Account); in the third simulation, we varied the BIAS parameter to examine expected patterns of translation equivalent (TE) learning as a function of word vocabulary (WORD) under the Avoidance, Preference, and Neutral Accounts. A summary of the parameter values used in each simulation is provided in Table 2.

@@

In Simulation 1, we examined patterns of translation equivalent learning where simulated children are at the same developmental level and thus have the same number of potentially LEARNABLE words, but their vocabulary size in each language (DOM and NONDOM) varied. For readers less familiar with this type of approach, we provide expanded examples of three simulated infants in the Appendix. For convenience, we set LEARNABLE constant at 600, which roughly corresponds to what is expected for an English--French 27-month-old bilingual child. For example, a 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and a 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages (Frank et al., 2016). BIAS was set constant at 1, meaning that in these simulations translation equivalents and singlets are equally easy to learn. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children's word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)).

In Figure 1, we plotted TE knowledge as a function of WORD, DOM, and NONDOM at different levels of BALANCE. Across all three Panels (1A, 1B, and 1C), simulated children with the most balanced vocabulary consistently produced more translation equivalents than other children. Moreover, Panels 1A and 1B show that, as WORD (word vocabulary) and DOM (dominant language words) increased, TE (translation equivalents) also increased regardless of BALANCE. Interestingly, Panel 1C shows that NONDOM and TE were extremely tightly coupled. In sum, we observed three important patterns, which served as Prediction Set 1 from the Bilingual Vocabulary Model for Study 2:

\begin{itemize}
\tightlist
\item
  Prediction 1a: Children with more balanced vocabularies (BALANCE) will produce more translation equivalents (TE) (Figure 1, Panel 1A).
\item
  Prediction 1b: Children who produce more total words (WORD) or more dominant-language words (DOM) will produce more translation equivalents (TE) (Figure 1, Panels 1A and 1B).\\
\item
  Prediction 1c: Children who produce more non-dominant language words (NONDOM) will produce more translation equivalents (TE). Unlike for WORD and DOM, this will not interact with BALANCE; instead, non-dominant vocabulary size will be an almost perfect predictor of translation equivalent knowledge (Figure 1, Panel 1C).
\end{itemize}

\begin{verbatim}
## Warning: Removed 20 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 140 rows containing missing values (geom_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 20 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 140 rows containing missing values (geom_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 3 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 25 rows containing missing values (geom_smooth).
\end{verbatim}

@@

In our previous simulation, we assumed that each simulated child was at the same developmental level and had the capacity to learn up to 600 words in each language (i.e., LEARNABLE was held constant at 600). As laid out in the introduction, under the Bilingual Vocabulary Model, the learnability of different words changes with a child's developmental level, where LEARNABLE increases as a child grows older. Therefore, in Simulation 2, we looked at the expected patterns of translation equivalent learning across varying levels of LEARNABLE (i.e., the number of learnable words in each language as developmental level changes). Additionally, we further examined vocabulary composition by computing the number of singlets in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) language. BIAS was once again kept constant at 1.

Translation equivalent knowledge was simulated across children at six developmental levels (number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of CONCEPT, TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will:

@@

\hypertarget{simulation-3-bias-towards-or-against-translation-equivalent-learning}{%
\subsection{1.3 Simulation 3: Bias towards or against translation equivalent learning}\label{simulation-3-bias-towards-or-against-translation-equivalent-learning}}
